### 2018-02-06 ident/docs/ident_manuscript.tex
copied text from committee meeting document

\section{Results:}
First in Section \ref{sec:closedform}, we show that the ability to establish practical identifiability of kinetic parameters in metabolic network models using our methodology, described in Section \ref{sec:ident}, relies upon the nonlinearity of the kinetic rate law formulations used to describe the fluxes. We use the gluconeogenic model described in Section \ref{sec:example} as an example. Then, in Section \ref{sec:trends} we show how the degree of identifiability of the maximum reaction rate parameter is always higher than the degree of identifiability of the corresponding enzyme binding parameter irrespective of enzyme kinetics rate law used to describe the flux. In Section \ref{sec:experiment_type} that follows, we discuss the the ability to determine experiments for parameter identifiability for a given flux based on the informativeness of a given experiment. In this section, we show how the informativeness of a given type of experiment to identify a specific flux can be deduced from its contribution towards the practical identifiability of the parameters for a given flux. In Section \ref{sec:noise} we discuss the identifiability of parameters when data with additive noise is used with our methodology to test practical identifiability. Finally, in Section \ref{sec:data_utility} we provide a motivation for the need for experimental design, especially to identify kinetic models metabolic networks.

\subsection{Nonlinearity of enzyme kinetic rate law expression affects identifiability analysis:}\label{sec:closedform}
Establishing parameter identifiability for fluxes in kinetic models of metabolism using the methodology we describe in the Methods section is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. We demonstrate one example of how the methodology works in Section \ref{sec:example} for parameters of flux $v_1$. The expression in Equation (\ref{eq:v1_par}) is obtained by using a computer algebra system. To recall, in this specific case, a computer algebra system is used to solve for the parameters of a flux described using the Michaelis-Menten kinetic rate law when experimental data from two different experiments is available.

@@

Using this expression for identifiability analysis, we find that each of the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ have two different close form expressions owing to the presence of a square root term in their solutions. These distinct expressions are denoted by (1) and (2) following the respective parameter names throughout the rest of the document: $V_3^{max}(1)$, $K_3^{fdp}(1)$, $K_3^{pep}(1)$, and $V_3^{max}(2)$, $K_3^{fdp}(2)$, $K_3^{pep}(2)$.

The impact of nonlinearity of a kinetic rate law expression affecting the practical identifiability of parameters in the expression may not be an unique problem isolated to the system that we are investigating. Complicated expressions for describing fluxes have been extensively used to model observed experimental data for different fluxes in a variety of organisms \parencite{Chassagnole2002a, Peskov2012, VanHeerden2014}. However, the identifiability of the parameters used in these models has never been truly examined. If these models are tractable, we believe our methodology can help in elucidating the identifiability of parameters in these models. If not, metabolic network fluxes can be expressed using alternative kinetic rate law models whose parameters can be tested for identifiability, and subsequently experiments can be designed for their estimation and model validation.

Next, we look some of the results we obtained for the practical identifiability of different parameters for all three fluxes $v_1$, $v_2$ and $v_3$ in the small gluconeogenic network model described in Section \ref{sec:small-model}, with the flux for $v_3$ described by Equation (\ref{eq:flux3_convkin}).

\subsection{Maximum reaction rates are more identifiable than enzyme binding constants:}\label{sec:trends}
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure1/v1_v2_v3_identifiability_no_noise}}
	\caption{The number of data combination from 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data is shown. The percentage of total experimental data combinations that can identify each parameter is specified in/on the right hand side of the bar showing the total number of combinations identifying a given parameter. The total number of data combinations used to identify parameters in fluxes $v_1$ and $v_2$ is lower that the total number of combinations used to test identifiability of parameters for $v_3$ since both $v_1$ and $v_2$ have only two parameters while $v_3$ is modeled with three parameters. Hence, the total number of experiments used to identify parameters in $v_1$ and $v_2$ is two, while parameter identification in $v_3$ requires data from at least three different experiments.}\label{fig:figure1}
\end{figure}
In Figure \ref{fig:figure1} we show the number and percentage of combinations that are capable of identifying each parameter in each flux. Based on the definition given in Section \ref{sec:degree_of_identifiability}, the percentages refer to the degree of identifiability of each parameter. The three panels in Figure \ref{fig:figure1} represent the degree of identifiability for parameters modeling the three different fluxes of the small network individually. Looking at these panels for $v_1$ (Figure \ref{fig:figure1}a), $v_2$ (Figure \ref{fig:figure1}b) and $v_3$ (Figure \ref{fig:figure1}c), the degree of identifiability of the maximum reaction rates ($V_i^{max}$) in each of the three fluxes is higher than the degree of identifiability of the enzyme binding ($K_i$) constants and the allosteric activation constant ($K_3^{pep}$). We observe this trend irrespective of the fact that only  $v_1$ and $v_2$ are represented by the same enzyme kinetic rate law (Michaelis-Menten), while the convenience kinetic rate law is used to model $v_3$. These observations lead us to conclude that the maximum reaction rate parameters are always more identifiable (as indicated by the higher degree of identifiability) than their enzyme binding constant counterparts irrespective of the enzyme kinetic rate law used to model the corresponding flux.

The most conspicuous of all observations from Figure \ref{fig:figure1} is the difference in the identifiability of parameters based on the flux in which these parameters are present. The variation in the degree of identifiability is not only dependent on the type of parameter being identified, but also depends on the  flux whose parameters are being tested for identifiability. This can be tied to selecting experimental data that have the ability to satisfy conditions for practical identifiability for a given parameter. We discuss the dependence of the variation in the degree of identifiability for parameters of different fluxes from the point of view of the informativeness of the experiment from which data is collected for identifiability in the following section.

\subsection{Experiments required for identifying parameters depends on the position of the flux in the metabolic network:}\label{sec:experiment_type}
In system identification, any input signal should be rich or informative enough to guarantee full excitement of the dynamics of the system. Only information obtained from such changes in the input and the corresponding signals can be used to completely identify the system over its entire dynamic range. Input signals that guarantee to fully excite the dynamics of the system are termed as persistently excitable signals. In linear systems, persistence of excitation of any input signal is guaranteed. However, for nonlinear systems, like the metabolic network we deal with in this paper, persistence of excitation of input signals is not guaranteed and should be assessed on a case by case basis. The lack of persistence of excitation requires the design of experiments to satisfy the data needs for complete system identification.

In this section, we demonstrate a specific case of lack of persistence of excitation and the subsequent experimental design required to overcome this hurdle for parameter identifiability for a small nonlinear metabolic network model. The informativeness of experiments for identifying parameters in a given flux depends on its ability to satisfy the conditions determined for practical identifiability of that parameter. For example, we demonstrated in Section \ref{sec:example} that for a combination of any two experiments to be capable of identifying $V_1^{max}$ and $K_1^{ac}$ in $v_1$, the experiments must have distinct acetate concentrations as well as a different uptake flux $v_1$ between them. Thus, in this instance, the informativeness of the experiments (changes in the input acetate concentration) for identifying parameters of $v_1$ is determined by their ability to change the measured value of flux $v_1$. Similarly, the identification of parameters for $v_2$ requires the experiments to distinguish between values of both $v_2$ as well as \textit{pep}, given both these variables are outputs of the model describing $v_2$.

In Figure \ref{fig:figure1}a we show that only about 20\% of all the experimental data sets satisfy these requirements for $v_1$ while more than 90\% of the available experimental satisfies these requires for $v_2$ (Figure \ref{fig:figure1}b). Thus, experimental design becomes crucial to determine the minimum number, as well as the nature of experiments that can help identify parameters for both $v_1$ and $v_2$. Recall that we use experimental data from five different experiments to test the practical identifiability of parameters in the model (Section \ref{sec:experiments}).
In order to determine the type of experiments that help satisfy identifiability conditions for both parameters of $v_1$, we look at the distribution of the five different experiments in all data combinations that can identify these 2 parameters (Figure \ref{fig:figure2}).

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v1max_k1ac_experiments}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_1^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination, and $K_1^{ac}$ c) experiment 1 ($k=1$), and d) experiment 2 ($k=2$) in combination. The percentages reflect the fractional contribution of each experiment type towards all identifiable data combinations.}\label{fig:figure2}
\end{figure}

The contribution of experiments that involve changes in the acetate concentrations, which consequently bring about changes in the value of $v_1$ contribute to a significant part (> 50\%) of the identifiable experimental data combinations in contrast to the other types of experiments. This matches with requirements for identifiability on the basis of the informativeness of experiments that was laid out earlier i.e., experiments must be able distinguish between changes in $v_1$ on the basis of changes in acetate. This informs why only about 20\% of the 306 different experiment combinations can identify these parameters (Figure \ref{fig:figure1}a). In can be gathered that only about 20\% of all the data combinations used to test identifiability are informative enough to distinguish between both acetate as well as $v_1$, and most of these experiments require changes in the acetate concentration to be able to identify $V_1^{max}$ and $K_1^{ac}$ through changes in the input and the output quantities acetate and $v_1$, respectively.

In contrast to $v_1$, we see that the different perturbation experiments have a higher contribution towards experimental data combinations that can identify parameters for $v_2$ (Figure). Also, it is important to note that since most experiment types can bring about changes in the dynamics of $v_2$ (changes to \textit{pep} and value of $v_2$), the identifiability of parameters for $v_2$ is relatively higher than the identifiability of $v_1$ parameters (Figure \ref{fig:figure1}b). We can further extend these arguments to also justify the observed contribution of experiments towards identifying parameters for $v_3$ (figure not shown).

Alternatively, both the degree of identifiability and the capability of different experiments to identify parameters of different fluxes based on their informativeness can be explained by the position of the flux in the metabolic network. We believe that the position of any given in the metabolic network determines the specific experiment that is persistently excitable enough to identify the parameters of that flux. This dependency of experiment informativeness on the position of the flux can be further elucidated using $v_1$ and $v_2$ as examples. We know from Equation (\ref{eq:v1_par}) that identifiability of $v_1$ requires changes in both acetate and $v_1$. We also know, based on our knowledge of the Michaelis-Menten kinetic rate law that changes in the substrate concentration of a reaction can bring about a nonlinear change in the value of the corresponding reaction rate (flux). In this specific metabolic network, since the substrate is an input variable to the model and $v_1$ is the corresponding uptake flux, the substrate can be easily perturbed to create persistently excitable experiments to identify parameters in $v_1$. We can generalize this observation for the identification of all uptake fluxes in all metabolic networks, i.e., at a minimum, a change in the input substrate concentration may be necessary for an informative experiment to identify the uptake flux parameters.

On the other hand, the Michaelis-Menten model for $v_2$ also requires changes in \textit{pep} and $v_2$ for persistently excitable experiments to identify $v_2$. However, since both of these are system outputs, satisfaction of this condition cannot be guaranteed without an analysis of the dynamics of the metabolic network and how changes in the input (acetate) bring about changes in the two requisite output quantities. Previously, a dynamical analysis of the current network has shown that the concentration of \textit{pep}, and hence the flux $v_2$ are functions of not only the input acetate concentration, but also the levels of expression of the different enzymes within the network (Srinivasan et al., 2017). Hence, it is theoretically possible for any of the five different experiment types to be persistently excitable to identify $v_2$. This is confirmed by the high degree of identifiability for both parameters of $v_2$, where in more than 90\% of all data combinations can identify the parameters.
Thus, this analysis informs us that the degree of identifiability and consequently, the type of experiments needed to identify different parameters varies widely depending the position of the flux in the metabolic network with respect to the inputs and the outputs of the network.

Nonetheless, all these explanations could only be verified due to the fact that the conditions for identifiability for both $v_1$ and $v_2$ are simple and readable due to the use of the Michaelis-Menten rate law to model these fluxes. Instead, in addition to presence of an allosteric interaction, when the convenience kinetic rate law is used to model this activation in $v_3$, we find that the complex expressions obtained for the three parameters preclude such an in-depth insight into the nature of persistently excitable experiments to identify  $v_3$. Thus, the only avenue through which experiments can be designed for the identification of parameters of $v_3$ would be through our proposed methodology wherein manual curation of experimental combinations required for identifiability are not necessary.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight, keepaspectratio]{C:/Users/shyam/Documents/Courses/CHE1125Project/Results/ident/figure2/v2max_k2pep_experiments}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter $V_2^{max}$ a) experiment 1 ($k=1$), and b) experiment 2 ($k=2$) in combination. Different experiments that contribute towards identifiability of $K_2^{pep}$ c) experiment 1 ($k=1$) and d) experiment 2 ($k=2$) in combination.}\label{fig:figure3}
\end{figure}

\subsection{Effect of noise on degree of identifiability is dependent on nonlinearity of enzyme kinetic rate law models:}\label{sec:noise}
The degree of identifiability for all the parameters shown in Figure \ref{fig:figure1} is obtained using noise-free in silico experimental data. However, actual experimental data from biological systems is usually noisy. Data gathered from experiments is usually associated with both biological noise attributable to the stochasticity of cellular function, as well as measurement noise associated with the techniques used to obtain concentration and flux measurements. In order to test the practical identifiability of parameters under circumstances where the presence of noise in the experimental data can affect the identifiability of the parameters, we apply our methodology on data (concentration and flux) with 5\% additive noise. 50 different samples of noise with 0 mean and 0.05 standard deviation are drawn from a normal distribution and added to the steady state experimental data generated on the basis of experiments described in Section \ref{sec:experiments} to generate 50 different samples of noisy data. Subsequently, all 50 samples are tested for their ability to practically identify parameters for all three fluxes of the small network. The corresponding graphs generated for practical identifiability analysis using noisy experimental data are shown in the Appendix.

@@

Despite the presence of noise, the degree of identifiability of $v_1$ and $v_2$ (figure not shown) does not change from the case where no noise is present in the experimental data (Figure \ref{fig:figure1}). However, the degree of identifiability for parameters of $v_3$ is affected by the presence of noise in the data (Figure). This effect is represented by the presence of a non-zero standard deviation and corresponding error bars on the graphs in the degrees of identifiability of parameters modeling $v_3$ (Figure). The standard deviation and the error bars represent the presence of data combinations that can identify a given parameter under certain values of noise, but not do so under different noise values. In the presence of 5\% additive noise in the data, the conditions under which parameters for $v_1$ and $v_2$ can be identified are strictly satisfied by the same data combinations that can identify these parameters in the absence of noise, i.e., the changes in the acetate concentration (for $v_1$) and \textit{pep} (for $v_2$) are significant enough to enable identifiability. The contrinution of different experiment types towards identifiability also remains the same (figure not shown separately).

@@

These observations, in conjunction with the estimation of the degree of identifiability of the different parameters (Figure \ref{fig:figure1}), helps in determining not only the nature of experiments required for parameter identifiability (Figure \ref{fig:figure2}), but also the order which the type of experiments necessary for identifying the entire metabolic network needs to be evaluated.

>>>>

\section{Results:}
First in section \ref{sec:closedform}, we show that the ability to establish practical identifiability of kinetic parameters in metabolic network models using our methodology, described in section \ref{sec:ident}, relies upon the nonlinearity of the kinetic rate law formulations used to describe the fluxes. We use the gluconeogenic model (Figure \ref{fig:network}) described in section \ref{sec:example} as an example. 
Then, in section \ref{sec:data_utility} we provide a motivation for the need for experimental design, especially to identify kinetic models of metabolic networks. We do this by looking at the number of combinations that can identify the maximum number of parameters. 
In section \ref{sec:trends} we show that the degree of identifiability of the maximum reaction rate parameter is always higher than the degree of identifiability of the corresponding enzyme binding parameter irrespective of enzyme kinetic rate law used to describe the flux. 
In section \ref{sec:experiment_type} we discuss the ability to determine the type of experiments for parameter identification for a given flux based on the informativeness of a given type of experiment. In this section we show how the informativeness of a given type of experiment to identify a specific flux can be deduced from its contribution towards the practical identification of the parameters for a given flux. Finally, we discuss some results arising out of the use of our methodology to determine the identifiability of parameters when data with additive noise is used in section \ref{sec:noise}.

\subsection{Nonlinearity of enzyme kinetic rate law expression affects identifiability analysis}\label{sec:closedform}
Practical identifiability of parameters in kinetic models of metabolism using the methodology described in section \ref{sec:methods} is governed by the nonlinear complexity of the enzyme kinetic rate law used to model a specific flux. We demonstrate one example of how the methodology works in section \ref{sec:example} for parameters of flux $v_1$. The expression in Equation (\ref{eq:v1_par}) is obtained by using a computer algebra system. To recall, in this specific case, a computer algebra system is used to solve for the parameters of a flux described using the Michaelis-Menten kinetic rate law when data from two different experiments is available.

@@

Using this expression for analysis, we find that each of the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ have two different closed-form expressions owing to the presence of a square root term in their solutions. These distinct expressions are denoted by (1) and (2) following the respective parameter names throughout the rest of the document: $V_3^{max}(1)$, $K_3^{fdp}(1)$, $K_3^{pep}(1)$, and $V_3^{max}(2)$, $K_3^{fdp}(2)$, $K_3^{pep}(2)$.

The impact of nonlinearity of a kinetic rate law expression affecting the practical identifiability of parameters in the expression may not be an unique problem isolated to the system that we are investigating. Complicated expressions for describing fluxes have been extensively used to model observed experimental data for different fluxes in a variety of organisms \parencite{Chassagnole2002a, Peskov2012, VanHeerden2014}. However, the identifiability of the parameters used in these models has never been truly examined. Unlike the case of the MWC model, if determining closed-form expressions for the parameters of these models are tractable, we believe our methodology can help in elucidating the identifiability of these parameters. If not, metabolic network fluxes can be expressed using alternative kinetic rate law models whose parameters can be tested for identifiability, and subsequently experiments can be designed for their estimation and model validation.

Next, we look at some of the results we obtained for the practical identification of different parameters for all three fluxes $v_1$, $v_2$ and $v_3$ in the small gluconeogenic network model described in section \ref{sec:small-model}, with the flux for $v_3$ described by Equation (\ref{eq:flux3_convkin}). We start by looking at one of motivations for the need for designing experiments to collect data for parameter estimation.

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{figures/figure4/v1_V1max_v2_v3_value1_data_utility}}
	\caption{Utility of experimental data combinations on the basis of their ability to identify the most number of parameters. Information is shown for parameters modeling fluxes a) $v_1$, b) $v_2$ and c) $v_3$. The total possible number of parameters that can be identified by data from combinations of a) two, b) two, and c) three experiments is shown on the horizontal axis. The vertical axis represents the total number of combinations that can identify the corresponding number of parameters. The percentages shown in the plots represent the fraction of the total combinations used to test identifiability of parameters for a given flux. A total of 306 data combinations are used for identifiability analysis for a) $v_1$ and b) $v_2$, and 4896 combinations are used to practical identifiability of c) $v_3$. Section \ref{sec:experiments} provides more details on how the combinations of experimental data are generated.}\label{fig:figure4}
\end{figure}	
\subsection{Utility of experimental data for practical parameter identification}\label{sec:data_utility}
Although the need for better parameter estimation methods is well researched, within the kinetic modeling community, the need for experimental design methods to satisfy the data needs for parameter estimation are not considered seriously. Hence, we find the need to stress the necessity for experimental design methodologies, in this case, tailored specifically for metabolic network models.

We do this by showing how useful the multiple combinations of experimental data are towards identifying parameters for each given flux (Figure \ref{fig:figure4}). We express the usefulness of any given data combination on the basis of the number of parameters that the combination can identify. The higher the number of parameters a data combination can identify, the greater is the usefulness of that combination of experiments.

The number of data combinations that can identify a specific number of parameters within a given flux is shown in Figure \ref{fig:figure4}. The percentage of all data combinations that can estimate a given number of parameters within a given flux are also shown in Figure \ref{fig:figure4}. For flux $v_1$ (Figure \ref{fig:figure4}a) about 78\% of the 306 data combinations cannot identify any parameter used to model the flux ($V_1^{max}$ and $K_1^{ac}(ne)$). Only a single parameter can be identified by about 2\% of the combinations, while only about 19\% of the data combinations can identify both the parameters used in the model. 	
In contrast, most data combinations (>90\%) are not wasted, and can identify both parameters in $v_2$ (Figure \ref{fig:figure4}b). Less than 5\% of the experimental data combinations are wasted from not being able to identify any parameter. In the case of $v_3$ (Figure \ref{fig:figure4}c), about 45\% of the 4896 data combinations can identify only one parameter, while 13\% of the combinations cannot identify any parameter. Only 10\% of combinations can identify $v_3$ completely.

Thus, Figure \ref{fig:figure4} gives credence to the idea that careful experimental design is necessary to minimize usage of resources devoted to performing experiments for parameter estimation. For instance, choosing any of the experiment combinations from the 78\% percent that does not identify any parameter in $v_1$, or the 13\% that does not identify any parameter in $v_3$ can lead to experiments that are not informative. This can lead to potential waste of resources used to perform these experiments.

In addition to providing information on the fraction of experiments that are informative and those that are not informative, Figure \ref{fig:figure4} also informs on the ease of identifying parameters for a single flux. For the small network (Figure \ref{fig:network}), we find that more than 90\% of all the data combinations tested for their ability to identify $v_2$ are actually capable of identifying $V_2^{max}$ and $K_2^{pep}$ (Figure \ref{fig:figure4}b). In contrast, less than 20\% of the combinations can identify both parameters in $v_1$. Since minimization of resource utilization requires us to use the least number of experiments to identify both $v_1$ as well as $v_2$, this indicates the need to carefully choose combinations of experiments that can identify $v_1$. A careful consideration of experiments for identifying $v_2$ is however not a major concern since there is a very high probability that experiments chosen to identify $v_1$ will also identify $v_2$.

Although the calculation of the utility of experiment combinations in terms of their ability to identify the maximum number of parameters gives an idea on the informativeness of experiments used to gather the data, further analysis from the perspective of identifiability of the parameters of different fluxes is required to be able to design experiments based on this information. Hence, we look at the degree of identifiability (see section \ref{sec:degree_of_identifiability}) of each parameter of each flux in the following section to be able to classify parameters, and consequently fluxes, from most identifiable to least identifiable.

\subsection{Maximum reaction rates are more identifiable than enzyme binding constants}\label{sec:trends}
\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.3\textheight]{figures/figure1/v1_V1max_v2_v3_identifiability}}
	\caption{The number of data combination from 18 different in silico experiments that can practically identify each parameter in fluxes a) $v_1$, b) $v_2$ and c) $v_3$ when there is no noise in the input experimental data. The percentage of total experimental data combinations that can identify each parameter is specified either in, or right next to the bar showing the total number of combinations identifying a given parameter. Since the total number of experiments used to identify parameters in $v_1$ and $v_2$ is two, the total number of data combinations used to identify parameters in fluxes $v_1$ and $v_2$ is lower (at 306) that the total number of combinations used to test identifiability of parameters for $v_3$ (at 4896), which is modeled with three parameters and requires data from at least three different experiments.}\label{fig:figure1}
\end{figure}
In Figure \ref{fig:figure1} we show the number and percentage of combinations that are capable of identifying each parameter in each flux. Based on the definition given in section \ref{sec:degree_of_identifiability}, the percentages refer to the degree of identifiability of each parameter. The three panels in Figure \ref{fig:figure1} represent the degree of identifiability for parameters modeling the three different fluxes of the small network individually.

In regards to the identifiability of $v_1$ (Figure \ref{fig:figure1}a), $v_2$ (Figure \ref{fig:figure1}b) and $v_3$ (Figure \ref{fig:figure1}c), the degree of identifiability of the maximum reaction rates ($V_i^{max}$) in each of the three fluxes is higher than the degree of identifiability of the corresponding enzyme binding ($K_i$) constants and the allosteric activation constant ($K_3^{pep}$). We observe this trend irrespective of the fact that only  $v_1$ and $v_2$ are represented by the same enzyme kinetic rate law (Michaelis-Menten), while the convenience kinetic rate law is used to model $v_3$. These observations lead us to conclude that the maximum reaction rate parameters are always more identifiable (as indicated by their higher degree of identifiability) than their enzyme binding constant counterparts, irrespective of the enzyme kinetic rate law used to model the corresponding flux.

From Figure \ref{fig:figure1} we see that the variation in the degree of identifiability is not only dependent on the type of parameter being identified, but is also associated with the flux that these parameters model.	In systems identification terminology, this can be tied to selecting experiments that are persistently excitable for the flux being identified. In systems identification, any input signal should be rich or informative enough to guarantee full excitement of the dynamics of the system \parencite{Ljung1994}. Only information obtained from such changes in the input can be used to completely identify the system over its entire dynamic range. Input signals that guarantee to fully excite the dynamics of the system are termed as persistently excitable signals. In linear systems, persistence of excitation of any input signal can be theoretically guaranteed. However, for nonlinear systems, like the metabolic network we deal with in this paper, persistence of excitation of input signals cannot be theoretically guaranteed, and should be assessed on a case by case basis. The lack of persistence of excitation requires the design of experiments to satisfy the data needs for complete system identification.

In the case of identifying fluxes within a metabolic network individually, each flux is not independent from the other fluxes that form the metabolic network. Each flux is connected to one or more fluxes through the metabolites that not only act as substrates and products, but also as allosteric activators and inhibitors (e.g., effect of \textit{pep} on $v_3$ in Figure \ref{fig:network}). Hence, careful consideration of experiments is a necessity so that the data acquired can satisfy conditions for practical identifiability for all parameters modeling a flux, and subsequently, all fluxes within a network. In the next section we discuss the dependence of the variation in the degree of identifiability for parameters of different fluxes from the point of view of the informativeness of experiments from which data is collected for identification.

\subsection{Experiment requirements for identifying parameters depend on the position of the flux in the metabolic network}\label{sec:experiment_type}	
In this section, we demonstrate a specific case of lack of persistence of excitation and the subsequent experimental design required to overcome this hurdle for parameter identifiability for a small nonlinear metabolic network model. Earlier, we mentioned that the informativeness of experiments for identifying parameters in a given flux depends on its ability to satisfy the conditions determined for practical identifiability of that parameter. For example, we demonstrated in section \ref{sec:example} that for a combination of any two experiments to be capable of identifying $V_1^{max}$ and $K_1^{ac}$ in $v_1$, the experiments must have distinct acetate concentrations as well as a different uptake flux $v_1$ between them (Equation \ref{eq:v1_par}). Thus, in this instance, the informativeness of the experiments (changes in the input acetate concentration and $v_1$) for identifying parameters of $v_1$ is determined by the ability of the input change to effect a change in the measured value of $v_1$. Similarly, the identification of parameters for $v_2$ requires the experiments to distinguish between values of both $v_2$ as well as \textit{pep}. %, given both these variables are outputs of the model describing $v_2$.

As such, as per Figure \ref{fig:figure1}a only 20\% of all data combinations can satisfy these requirements, and can consequently identify $v_1$. Whereas, more than 90\% of the available data combinations satisfy the requirements for identifying $v_2$ (Figure \ref{fig:figure1}b). Thus, identifiability analysis is crucial to determine the minimum number of experiments, as well as the nature of experiments that can help identify parameters for both $v_1$ and $v_2$. Recall that we use experimental data from five different types of experiments to test the practical identifiability of parameters in the model (Section \ref{sec:experiments}).
In order to determine the type of experiments that help satisfy identifiability conditions for both parameters of $v_1$, we look at the distribution of the five different types in all data combinations that can identify these 2 parameters (Figure \ref{fig:figure2}a and b).

\begin{figure}[!tbhp]
	\centering{\includegraphics[width=1.0\textwidth,height=0.6\textheight, keepaspectratio]{figures/figure2/v1_V1max_only_v2_spider}}
	\caption{The contribution of different experiments types used in a combination of two experiments ($k = \{1, 2\}$) that can practically identify parameter a) $V_1^{max}$, b) $K_1^{ac}$, c) $V_2^{max}$ and d) $K_2^{pep}$. The percentages reflect the fractional contribution of each experiment type towards all identifiable data combinations.}\label{fig:figure2}
\end{figure}

The contribution of experiments that involve changes in the acetate concentrations, which consequently bring about changes in the value of $v_1$, contribute to a significant part (> 50\%) of the identifiable experimental data combinations in comparison to the other types of experiments. This matches with requirements for identifiability on the basis of the informativeness of experiments that was laid out earlier i.e., experiments must be able distinguish between different values of $v_1$ and acetate. Accordingly, only about 20\% of the 306 different experiment combinations used to test identifiability are informative enough to discriminate both acetate as well as $v_1$, and can identify $V_1^{max}$ and $K_1^{ac}(ne)$ (Figure \ref{fig:figure1}a).

In contrast to $v_1$, we see that the different enzyme perturbation experiments have a higher contribution towards data combinations that can identify parameters for $v_2$ (Figure \ref{fig:figure2}c and d). Hence, in conjunction with the high degree of identifiability seen in Figure \ref{fig:figure1}b for $v_2$, we can deduce that almost all experiment types are persistently excitable for $v_2$ i.e., most experiments can bring about noticeable changes to both \textit{pep} and $v_2$. Consequently, in comparison to selecting experiments to identify $v_1$, there is very little restriction on the types of experiments that are informative enough to identify $v_2$. This observations is in line with the one made in section \ref{sec:data_utility} based on the total number of combinations that can identify both parameters in $v_2$.

Both the degree of identifiability of parameters and the informativeness of the corresponding experiments used to identify them can be explained by the position of the flux in the metabolic network. The position of any given flux in the metabolic network determines the specific experiment that is persistently excitable enough to identify the parameters of that flux. This dependency of experiment informativeness on the position of the flux can be further elucidated using $v_1$ and $v_2$ as examples. We know from Equation (\ref{eq:v1_par}) that identifiability of $v_1$ requires changes in both acetate and $v_1$. We also know, based on our knowledge of the Michaelis-Menten kinetic rate law that changes in the substrate concentration of a reaction can bring about a nonlinear change in the value of the corresponding reaction rate. In this specific metabolic network, since the substrate is an input variable to the model, and $v_1$ is the corresponding uptake flux, the substrate can be easily perturbed to create persistently excitable experiments to identify parameters in $v_1$. We can generalize this observation for the identification of all uptake fluxes in all metabolic networks, i.e., at a minimum, a change in the input substrate concentration may be necessary for an informative experiment to identify the uptake flux parameters.

On the other hand, the Michaelis-Menten model for $v_2$ also requires changes in \textit{pep} and $v_2$ for persistently excitable experiments to identify $v_2$. However, since both of these are system outputs, satisfaction of this condition cannot be guaranteed without an analysis of the dynamics of the metabolic network and how changes in the input (acetate) bring about changes in the two requisite output quantities. Previous dynamical analysis of the network (Figure \ref{fig:network} in the Appendix) has already established the existence of a functional relationship between \textit{pep} and $v_2$, and the input acetate concentration and the levels of expression of the different enzymes within the network \parencite{Srinivasan2017}. Hence, it is theoretically possible for any of the five different experiment types to be persistently excitable to identify $v_2$. This is confirmed by the high degree of identifiability for both parameters of $v_2$, where in more than 90\% of all data combinations can identify the parameters. Thus, this analysis informs us that the degree of identifiability and consequently, the type of experiments needed to identify different parameters varies widely depending on the position of the flux in the metabolic network with respect to the inputs and the outputs of the network, as well as the various regulatory interactions present within the network. We can extend these observations to justify the observed contribution of experiments towards identifying parameters for $v_3$ as well (Figure \ref{fig:figure2_v3}).

\subsection{Effect of noise on degree of identifiability is dependent on nonlinearity of enzyme kinetic rate law models}\label{sec:noise}
In reality, experimental data from biological systems is usually noisy. Noise in data gathered from experiments is associated with biological noise attributable to the stochasticity of cellular function, as well as measurement noise associated with the techniques used to obtain concentration and flux measurements. However, the previous sections describe practical identifiability for parameters obtained using noise-free in silico experimental data. So, in order to illustrate the ability of our method to test the practical identifiability of parameters under realistic circumstances, we apply our methodology on in silico data (concentration and flux) with 5\% additive noise.

50 different samples of noise with 0 mean and 5\% standard deviation are drawn from a normal distribution and added to the steady state experimental data generated on the basis of experiments described in Section \ref{sec:experiments} to generate 50 different samples of noisy data. Subsequently, all 50 samples are tested for their ability to practically identify all three fluxes of the small network (Figure \ref{fig:network}). The degree of identifiability of the different parameters using noisy experimental data are shown in the Appendix (Figure \ref{fig:figure1_noisy}).

@@

Despite the presence of noise, the degree of identifiability of $v_1$ and $v_2$ (Figure \ref{fig:figure1_noisy}a and b in the Appendix) does not change from the case where no noise is present in the experimental data (Figure \ref{fig:figure1}). However, the degree of identifiability for parameters of $v_3$ is affected by the presence of noise in the data (Figure \ref{fig:figure1_noisy}c in the Appendix). This effect is represented by the presence of a non-zero standard deviation and corresponding error bars on the graphs in the degrees of identifiability of parameters modeling $v_3$ (Figure \ref{fig:figure1_noisy}c in the Appendix). The standard deviation and the error bars represent the presence of data combinations that can identify a given parameter under certain values of noise, but not do so under different noise values. In the presence of 5\% additive noise in the data, the conditions under which parameters for $v_1$ and $v_2$ can be identified are strictly satisfied by the same data combinations that can identify these parameters in the absence of noise, i.e., the changes in the acetate concentration (for $v_1$) and \textit{pep} (for $v_2$) are significant enough to enable identification. The contribution of different experiment types towards identifiability also remains the same (figure not shown separately).

@@

These observations, in conjunction with the estimation of the degree of identifiability of the different parameters (Figure \ref{fig:figure1}) help in determining not only the nature of experiments required for parameter identifiability (Figure \ref{fig:figure2}), but also the order in which the type of experiments necessary for identifying the entire metabolic network needs to be evaluated.
